Analysis of observational data, to calculate the Delayed-Tau model parameters

### 2025-02-09 thesis.qmd
mcmc theory

[^1]: @hoskinGreatDebateWhat1976

@@

@@

-   **UNGC Catalog** [@karachentsevUPDATEDNEARBYGALAXY2013; @karachentsevSTARFORMATIONPROPERTIES2013]**:** The Updated Nearby Galaxy Catalog (UNGC)[^5] provides a distance-limited sample of galaxies, within 11 Mpc or radial velocities of $V < 600\ km \cdot s^{−1}$. Selecting galaxies with $D\leq 11$ Mpc, we get a sample of 1321 galaxies. It offers detailed star formation rate estimates derived from H$\alpha$ and far-ultraviolet (FUV) fluxes, making it particularly valuable for studying nearby galaxies and local star formation processes. The UNGC catalog focuses on galaxies with well-determined distances and includes detailed environmental and morphological data.

[^5]: The UNGC tables can be downloaded using the following link: <https://www.sao.ru/lv/lvgdb/introduction.php>. Here, we use the latest update from 17.01.2025.

@@

-   **Luminosity and Distance Dependence:** The HECATE catalog is more complete for brighter galaxies (with luminosities above $10^{9.5} L_{B,\odot}$) and distances within 33 Mpc, which includes the LV. For galaxies with luminosities around $10^{10} L_{B,\odot}$)​, the catalog remains complete up to distances of 100 Mpc.
-   **Luminosity Density:** The catalog's completeness is also estimated based on the luminosity density in the B-band. Within 30 Mpc, the catalog is over 100% complete due to the high density of galaxies around the Milky Way. However, at distances around 170 Mpc, the completeness drops to approximately 50%, indicating that many galaxies are missing.
-   **Star Formation Rate and Stellar Mass:** The completeness of SFR is around 50% for distances between 30 and 150 Mpc, primarily due to limitations in the WISE survey coverage. Similarly, the completeness of stellar mass estimates is comparable to that of the B-band luminosity, with overdensities at short distances and a sharp cutoff at larger distances.

@@

The incompleteness for ultra-faint galaxies is primarily due to their low surface brightness, making them difficult to detect beyond the Local Group. This limitation is particularly significant for studies of dwarf galaxies and the faint end of the galaxy luminosity function

@@

For the comparison of our data we will use linear regression, accounting for the errors of the observed data, where $R^2$ and the slope of the fitted line should be close to 1[^6]. We will also plot the residuals ($y_\text{Regression}-y_\text{observed}$) and if the mean value $\overline{\text{Residuals}}\approx 0$, then the linear regression is considered succesfull.

[^6]: $R^2$ measures the proportion of variance explained by the linear model. For a model to be acceptable the $R^2$ should be higher than $80\%$, or close to it if $\overline{\text{Residuals}}\approx 0$

All the results are shown in @sec-regressions, except the comparisons of the Morphological Types and of the SFRs, that need more detailed analysis.[^7]

[^7]: The names of the x and y axes are written as they are given in the corresponding catalog, except the ones in decimal logarithmic scale, that are notated appropriately.

>>>>

[^1]: The chain is homogeneous if $T \forall x^{(i)}$ remains invariant (the probability of moving from $x^{(i−1)}$ to $x^{(i)}$ does not depend on the iteration number $i$.) and $\sum_{x^{(i)}} T\bigl(x^{(j)} \mid x^{(i-1)}\bigr) = 1$.

@@

*MCMC* combines the properties of Markov chains and Monte Carlo, allowing the approximation of aspects of posterior distributions that cannot be directly calculated.

Let $x^{(i)}$ be a homoguneous Markov chain[^5] inside a high-dimensional space $X$ of the set of possible configurations of a system and the space on which the posterior is defined, $x^{(i)}\in X = {x_1,x_2,...}$.

[^5]: The chain is homogeneous if $T \forall x^{(i)}$ remains invariant (the probability of moving from $x^{(i−1)}$ to $x^{(i)}$ does not depend on the iteration number $i$.) and $\sum_{x^{(i)}} T\bigl(x^{(j)} \mid x^{(i-1)}\bigr) = 1$.

The chain is described by the formula:

$$
p\bigl(x^{(i)} \mid x^{(i-1)},\dots,x^{(1)}\bigr) = T\bigl(x^{(i)} \mid x^{(i-1)}\bigr),
$$

where $T$ is the transition matrix and $p$ is the probability, following Bayes' rule $p(x|D) \propto p(D|x)\cdot p(x)$ ($D$ indicates the data). We chose $T$ in such a way that its stationary distribution is the posterior $p$:

$$
p\bigl(x^{(i)}\bigr)
\;=\;
\sum_{x^{(i-1)}} 
p\bigl(x^{(i-1)}\bigr)\;T\bigl(x^{(i)} \mid x^{(i-1)}\bigr).
$$

$T$ should be *irreducible*, so any state of the chain has a positive probability of visiting all the other states, meaning that $T$ cannot be broken into smaller tables, and *aperiodic*, so the chain does not get trapped in cycles.

In continuous state spaces, the transition matrix $T$ becomes an integral kernel $K$ and $p(x)$ becomes the corresponding eigenfunction

$$
 p\bigl(x^{(i)}\bigr) = \int p\bigl(x^{(i-1)}\bigr)\,K\bigl(x^{(i)} \mid x^{(i-1)}\bigr)\,dx^{(i-1)} .
$$

$K$ is the conditional density of $x^{(i)}$, and it is chosen by the corresponding MCMC algorithm, such as the Metropolis-Hastings algorithm.

In practice, this means that given a prior distribution for the parameters of the examed model (in our case we assume priors for the parameters $t_{sf},\ \tau,\ A_{del}$ from the delayed-$\tau$ model) and the likelihood based on observed data (we have experimental data for the SFR and $M_*$), a MCMC algorithm can calculate the posterior probability of the model and thus approximate the distribution of the parameters. From these distributions, their mean values are the values of the parameters and their variance is the error.

To calculate these, the algorithm firstly initializes some chains in the parameter space ($t_{sf},\ \tau,\ A_{del}$) and begins sampling the posterior distribution based on the values of the parameters. The first iterations are discarded, since if the sampling begins in a low probability region it can significantly affect the variance of results (burn-in period)[@andrieuIntroductionMCMCMachine2003]. Eventually, the chains should converge. To examine this we use the diagnostics $\hat{R}<1.01$ and ESS\>100. The diagnostics indicate if different chains have converged to the same posterior region and if the sample of draws autocorrelates, respectively.[^6]

[^6]: For more information about the metrics used by the Stan software you can visit <https://mc-stan.org/docs/2_36/reference-manual/analysis.html#effective-sample-size.section>

@@

-   **UNGC Catalog** [@karachentsevUPDATEDNEARBYGALAXY2013; @karachentsevSTARFORMATIONPROPERTIES2013]**:** The Updated Nearby Galaxy Catalog (UNGC)[^7] provides a distance-limited sample of galaxies, within 11 Mpc or radial velocities of $V < 600\ km \cdot s^{−1}$. Selecting galaxies with $D\leq 11$ Mpc, we get a sample of 1321 galaxies. It offers detailed star formation rate estimates derived from H$\alpha$ and far-ultraviolet (FUV) fluxes, making it particularly valuable for studying nearby galaxies and local star formation processes. The UNGC catalog focuses on galaxies with well-determined distances and includes detailed environmental and morphological data.

[^7]: The UNGC tables can be downloaded using the following link: <https://www.sao.ru/lv/lvgdb/introduction.php>. Here, we use the latest update from 17.01.2025.

@@

-   **Luminosity and Distance Dependence:** The HECATE catalog is more complete for brighter galaxies (with luminosities above $10^{9.5} L_{B,\odot}$) and distances within 33 Mpc, which includes the LV. For galaxies with luminosities around $10^{10} L_{B,\odot}$, the catalog remains complete up to distances of 100 Mpc.
-   **Luminosity Density:** The catalog's completeness is also estimated based on the luminosity density in the B-band. Within 30 Mpc, the catalog is over 100% complete due to the high density of galaxies around the Milky Way. However, at distances around 170 Mpc, the completeness drops to approximately 50%, indicating that many galaxies are missing.
-   **Star Formation Rate and Stellar Mass:** The completeness of SFR is around 50% for distances between 30 and 150 Mpc, primarily due to limitations in the WISE survey coverage. Similarly, the completeness of stellar mass estimates is comparable to that of the B-band luminosity, with overdensities at short distances and a sharp cutoff at larger distances.

@@

The incompleteness for ultra-faint galaxies is primarily due to their low surface brightness, making them difficult to detect beyond the Local Group. This limitation is particularly significant for studies of dwarf galaxies and the faint end of the galaxy luminosity function.

@@

For the comparison of our data we will use linear regression, accounting for the errors of the observed data, where $R^2$ and the slope of the fitted line should be close to 1[^8]. We will also plot the residuals ($y_\text{Regression}-y_\text{observed}$) and if the mean value $\overline{\text{Residuals}}\approx 0$, then the linear regression is considered succesfull.

[^8]: $R^2$ measures the proportion of variance explained by the linear model. For a model to be acceptable the $R^2$ should be higher than $80\%$, or close to it if $\overline{\text{Residuals}}\approx 0$

All the results are shown in @sec-regressions, except the comparisons of the Morphological Types and of the SFRs, that need more detailed analysis.[^9]

[^9]: The names of the x and y axes are written as they are given in the corresponding catalog, except the ones in decimal logarithmic scale, that are notated appropriately.
